---
title: "Model Selection"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
# Importing libraries
```{r}
install.packages('ISLR')
install.packages('leaps')
```

# Subset Selection Methods
## Best subset selection
```{r}
library(ISLR)
library(leaps)
attach(Hitters)

Hitters = na.omit(Hitters)
```

The regsubsets() function (part of the leaps library) performs best subset selection by identifying the best model that contains a given number of predictors, where best is quantified using RSS. The syntax is the same as for lm().

```{r}
regfit.full = regsubsets(Salary~., Hitters)
summary(regfit.full)
```

By default, regsubsets() only reports results up to the best eight-variable model. But the nvmax option can be used in order to return as many variables as are desired. Here we fit up to a 19-variable model.

```{r}
regfit.full = regsubsets(Salary~., data=Hitters, nvmax=19)
reg.summary = summary(regfit.full)
names(reg.summary)
reg.summary$rsq
```
Plotting RSS, adjusted R2, Cp, and BIC for all of the models at once will help us decide which model to select. Note the type="l" option tells R to connect the plotted points with lines.

```{r}
par(mfrow=c(2,2))
plot(reg.summary$rss, xlab="Number of Variables ", ylab="RSS", type="l")

plot(reg.summary$adjr2, xlab="Number of Variables ", ylab="Adjusted RSq", type="l")
max.adjr2 = which.max(reg.summary$adjr2)
points(max.adjr2, reg.summary$adjr2[max.adjr2], col="red", cex=2, pch =20)

plot(reg.summary$cp, xlab="Number of Variables ", ylab="Cp", type="l")
min.cp = which.min(reg.summary$cp)
points(min.cp ,reg.summary$cp[min.cp], col ="red", cex=2, pch =20)

plot(reg.summary$bic, xlab="Number of Variables ", ylab="BIC", type="l")
min.bic = which.min(reg.summary$bic)
points(min.bic, reg.summary$bic[min.bic],col="red",cex=2,pch =20)
```
The regsubsets() function has a built-in plot() command which can be used to display the selected variables for the best model with a given number of predictors, ranked according to the BIC, Cp, adjusted R2, or AIC.

```{r}
plot(regfit.full, scale="r2")
plot(regfit.full, scale="adjr2")
plot(regfit.full, scale="Cp")
plot(regfit.full, scale="bic")
```

The top row of each plot contains a black square for each variable selected according to the optimal model associated with that statistic. For instance, we see that several models share a BIC close to −150. However, the model with the lowest BIC is the six-variable model that contains only AtBat, Hits, Walks, CRBI, DivisionW, and PutOuts. We can use the coef() function to see the coefficient estimates associated with this model.

```{r}
coef(regfit.full, 6)
```

## Forward and Backward Stepwise Selection