---
title: "Missing Data"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---

# Getting Set Up

## Setting chunk options and generating R script
```{r results="hide"}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
knitr::purl("missing-data.Rmd")
```

## Installing Packages
```{r}
install.packages('mice')
install.packages('lattice')
install.packages('VIM')
install.packages('aod')
install.packages('BaM')

library(mice)
library(lattice)
library(VIM)
library(aod)
library(BaM)
```

## Loading in data
```{r}
titanic = read.table("titanic.txt", header=T, sep=",")
titanic.missing = titanic[,c(1,2,3,5,11)]
head(titanic.missing, 10)
```

# Exploratory Analysis
## Exploring the missingness using VIM
```{r}
titanic.missing.aggr = aggr(titanic.missing, numbers=TRUE, prop=FALSE, ylab=c("Histogram of missing data","Pattern"))
titanic.missing.aggr

aggr(titanic.missing, combined=TRUE, numbers=TRUE, prop=TRUE, cex.numbers=0.87, varheight=FALSE)
```
Amount of missingness in age for each survived group.
```{r}
barMiss(titanic.missing[,c("survived","age")])
```

Amount of missigness in age for each sex group
```{r}
barMiss(titanic.missing[,c("sex","age")])
histMiss(titanic.missing)
```

# Complete Cases
## Fitting a logistic regression model for the complete cases
```{r}
titanic.logistic.omit = glm(survived ~ pclass+sex+age, family=binomial, data=titanic.missing)
summary(titanic.logistic.omit)
```

## Global effect of class
```{r}
wald.test(b=coef(titanic.logistic.omit), Sigma=vcov(titanic.logistic.omit), Terms=2:3)
```

## Odds ratios
```{r}
exp(cbind(OR=titanic.logistic.omit$coefficients, confint(titanic.logistic.omit)))
```
 
# Multiple Imputation
## Studying the patterns of missingness
```{r}
pattern = md.pattern(titanic.missing)
pattern

pairs = md.pairs(titanic.missing)
pairs
```

## Imputing the missing values
```{r results="hide"}
imp = mice(titanic.missing, m=100)
imp
```

## Imputed values for age.
Each row corresponds to a missing entry in age. The columns contain the multiple imputations.
```{r}
imp$imp$age[1:10,1:5]
```
## Combine observed and imputed data.
Only first ten passengers shown.
```{r}
complete(imp,1)[1:10,]
```



## Diagnostic Checking
It is often useful to inspect the distributions of original and the imputed data. The complete() function extracts the original and the imputed data sets from the imp object as a long (row-stacked) matrix. The col vector separates the observed (blue) and imputed (red) data for age.
```{r}
com = complete(imp, "long", inc=T)
col = rep(c("blue","red")[1+as.numeric(is.na(imp$data$age))],101)
stripplot(age~.imp, data=com, jit=TRUE, fac=0.8, col=col, pch=20, cex=1.4, xlab="Imputation number")
```

# Analyses of Imputed Data
```{r}
# Analyzing the imputed data sets
fit = with(data=imp, exp=glm(survived ~ pclass + sex + age, family=binomial))

# Creating a data set with the results of all the analysis
MI.matrix<-matrix(0,100,5)
for(k in 1:100) {
  MI.matrix[k,] = coefficients(fit$analyses[[k]])
  MI.results = data.frame(Intercept=MI.matrix[,1], pclass2=MI.matrix[,2], pclass3=MI.matrix[,3], sex=MI.matrix[,4], age=MI.matrix[,5])
}

MI.results[1:10,]
```

## Combining the results using Rubin’s rule
The column fmi contains the fraction of missing information, i.e. the proportion of the variability that is attributable to the uncertainty caused by the missing data.
```{r}
est = pool(fit)
summary(est)
```

# Inverse Probability Weighting
## Creating the missing data indicator variable r
```{r}
titanic.missing$r = as.numeric(!is.na(titanic.missing$age))*as.numeric(!is.na(titanic.missing$sex))
head(titanic.missing,15)
```

## Fitting the logistic regression model to calculate the probabilities of being complete
```{r}
titanic.ipw.glm<-glm(r ~ pclass + survived, data=titanic.missing,family=binomial)
summary(titanic.ipw.glm)
```

## Calculating the weights: Inverse Probabilities
```{r}
titanic.missing$w<-1/fitted(titanic.ipw.glm)
head(titanic.missing, 15)
```

## Generating Model
```{r}
titanic.results.ipw = glm(survived ~ pclass + sex + age, data=titanic.missing, weights=titanic.missing$w, family=binomial)
summary(titanic.results.ipw)
```

