---
title: "Linear Regression"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---

---

# Importing Libraries
```{r}
install.packages('pastecs')
```

# Basics
Reading the data in R.
```{r}
library(pastecs)
kalama = read.table("kalama.txt", header=T)
attach(kalama)
kalama
```

Descriptive statistics in R.
```{r}
options(digits=2)
descrip.kalama = stat.desc(kalama[,c("age","height")],basic=TRUE, desc=TRUE)
descrip.kalama
```

Estimating Correlations in R.
```{r}
cov.age.height = cov(age, height)
corr.age.height = cor(age, height)
cov.age.height
corr.age.height
```

Testing if the population correlation is zero.
```{r}
corr.age.height.test = cor.test(age, height, alternative="two.sided", method="pearson")
corr.age.height.test
```

Scatterplot with line.
```{r}
plot(age, height, main="Age vs Height", xlab="Age", ylab="Height", pch=19)
abline(lm(height~age), col="red")
```

# Simple Linear Regression
```{r}
res = lm(height~age, data=kalama)
kalama.anova = anova(res)
kalama.summary = summary(res)
kalama.anova
kalama.summary
```

# Multiple Linear Regression

## Reading in the data
```{r}
satisfaction = read.table("satisfaction.txt", header=T)
attach(satisfaction)
satisfaction
```
## Exploring the data
```{r}
cor(satisfaction)
plot(satisfaction)
```

Descriptive statistics
```{r}
options(digits=2)
descrip.satisfaction = stat.desc(satisfaction,basic=TRUE, desc=TRUE)
descrip.satisfaction
```

## Fitting the model
```{r}
satisfaction.lm = lm(satis~age+severity+anxiety, data=satisfaction)
satisfaction.summary = summary(satisfaction.lm)
satisfaction.summary
```

## Likelihood ratio test null model versus full model
```{r}
satisfaction.lm.int = lm(satis~1, data=satisfaction) # Null model
anova(satisfaction.lm.int,satisfaction.lm) # Null versus full
```

## Sequential building of the model
```{r}
satisfaction.anova = anova(satisfaction.lm)
satisfaction.anova
```

## Final Model
```{r}
satisfaction.lm.final = lm(satis~age+anxiety, data=satisfaction)
satisfaction.final.summary = summary(satisfaction.lm.final)
satisfaction.final.summary
```

## Predicting a new observation
```{r}
newdata = data.frame(age=43, anxiety=2.7)
pred.w.plim = predict(satisfaction.lm.final, newdata, interval="predict")
pred.w.clim = predict(satisfaction.lm.final, newdata, interval = "confidence")
pred.w.plim
pred.w.clim
```