---
title: "Models for Longitudinal Data"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---

# Getting Set Up

## Setting chunk options
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

## Installing Packages
```{r}
install.packages('lattice')
library('lattice')
```

## Reading in the data
```{r}
early.int1 = read.table("earlyint.txt", header=T, sep=",")
attach(early.int1)
```

## Exploring the data

Spaghetti plot.
```{r}
n = length(unique(id))
interaction.plot(age, id, cog, xlab="Age in years", ylab="IQ", legend=F)
```
Descriptives.
```{r}
early.mean=tapply(cog,list(age,program),mean) #mean
early.sd=tapply(cog,list(age,program),sd) #sd
early.var=tapply(cog,list(age,program),var) #variance
early.n=table(age,program) #frequency
```

Boxplots.
```{r}
boxplot(cog~age, xlab="Age (in years)", ylab="IQ")
```

Boxplots per program.
```{r}
par(mfrow=c(2,1))
boxplot(cog[program==0]~age[program==0], main="No intervention", main="No intervention", xlab="Age (in years)", ylab="IQ")
boxplot(cog[program==1]~age[program==1], main="Intervention", main="No intervention", xlab="Age (in years)", ylab="IQ")
```

General function to plot error bars.
```{r}
errbar=function(x,y,height,width,lty=1,col="black"){
arrows(x,y,x,y+height,angle=90,length=width,lty=lty, col=col)
arrows(x,y,x,y-height,angle=90,length=width,lty=lty, col=col)
}
```


Plotting mean evolutions.
```{r}
plot(age[id==1],early.mean[,1],type="b",xlim=c(1,2), ylim=c(40,160),xlab="Age (in years)",ylab="IQ",axes=F, main="Mean evolution (with 1 SE intervals)")
axis(side=1,at=c(1,1.5,2),labels=c(1,1.5,2))
axis(side=2,at=seq(40,160,20))

box()
points(age[id==1],early.mean[,2],type="b",col="red")
errbar(age[id==1]-.005,early.mean[,1],early.sd[,1],.1)
errbar(age[id==1]+.005,early.mean[,2],early.sd[,2],.1,col="red")
```

Correlation between IQ scores at different ages.
```{r}
early.int2 <- reshape(early.int1, timevar = "age", idvar = c("id", "program"), direction = "wide") # reshaping data into wide form
early.int2

 cor(early.int2[,3:5])
```

# Linear Regression Per Person
Creating the time variable.
```{r}
early.int1$age0<-early.int1$age-1
```

Displaying the linear regression per person.
```{r}
cf = sapply(early.int1$id, function(x) coef(lm(cog~age0, data=subset(early.int1, id==x))))
Sx<-reorder(early.int1$id, cf[1,])

xyplot(cog ~ age0|Sx, groups=program, data=early.int1, type=c('p','r'), auto.key=T,aspect="xy", par.settings=list(axis.text=list(cex=0.6), fontsize=list(text=8, points=10)), scales=list(x=list(at=c(0,0.5,1),labels=c("0","0.5","1")))
)
```

## Linear regression of cog on age per participant.

Coefficients.
```{r}
lin.reg.coef = by(early.int1, early.int1$id, function(data) coef(lm(cog ~ age0, data=data)))
lin.reg.coef1 = unlist(lin.reg.coef)
names(lin.reg.coef1) = NULL
lin.reg.coef2 = matrix(lin.reg.coef1,length(lin.reg.coef1)/2,2,byrow = TRUE)
```

R-Squared.
```{r}
lin.reg.r.squared = by(early.int1, early.int1$id, function(data) summary(lm(cog ~ age, data=data))$r.squared)
lin.reg.r.squared1 = as.vector(unlist(lin.reg.r.squared))
```

Histograms.
```{r}
par(mfrow=c(3,1))
hist(lin.reg.coef2[,1],xlab="Intercept",col="lightblue",main="Histogram of individual intercepts")
hist(lin.reg.coef2[,2],xlab="Slope",col="lightblue",main="Histogram of individual slopes")
hist(lin.reg.r.squared1,xlab="R squared",col="lightblue",main="Histogram of individual R squared")
```

# Linear regression per person and group

Plotting individual regression lines per group.
```{r}
reg.coef = cbind(lin.reg.coef2, early.int1[early.int1$age==1,]$program)
mean.int = tapply(reg.coef[,1], reg.coef[,3], mean)
mean.slope = tapply(reg.coef[,2], reg.coef[,3], mean)

par(mfrow=c(1,2))
plot(age, cog, type="n", xlim=c(1,2), ylim=c(40,160), main="No intervention", xlab="Age-1 (in years)", ylab="IQ", axes=F)
axis(side=1, at=c(1,1.5,2), labels=c(1,1.5,2))
axis(side=2, at=seq(40,160,20))
box()

for (i in 1:103) {
  if (reg.coef[i,3]==0){ 
    curve(cbind(1,x)%*%reg.coef[i,1:2], add=T, col="gray") 
    curve(cbind(1,x)%*%c(mean.int[1], mean.slope[1]), add=T, lwd=2)
  }
}

plot(age, cog, type="n", xlim=c(1,2), ylim=c(40,160),main="Intervention", xlab="Age-1 (in years)", ylab="IQ", axes=F)
axis(side=1, at=c(1,1.5,2), labels=c(1,1.5,2))
axis(side=2, at=seq(40,160,20))
box()

for (i in 1:103){
  if (reg.coef[i,3]==1){
    curve(cbind(1,x)%*%reg.coef[i,1:2], add=T, col="gray") 
    curve(cbind(1,x)%*%c(mean.int[2], mean.slope[2]), add=T, lwd=2)
  }
}
```

# Fitting the Model

## Installing the Packages
```{r}
install.packages("lme4")
install.packages("arm")
install.packages("pbkrtest")

library(lme4)
library(lattice)
library(arm)
library(car)
library(pbkrtest)
```

Creating the time variable.
```{r}
early.int1$age0<-early.int1$age-1
```

Fitting the model with maximum likelihood.
```{r}
early.lmer1 = lmer(cog~1+age0*program+(1 + age0|id), REML = FALSE, data=early.int1)
summary(early.lmer1)
```

## Estimating the fixed effects via bootstrap
```{r}
fixed.boot = bootMer(early.lmer1, fixef, use.u = TRUE, nsim = 250)
fixed.boot
summary(fixed.boot)
```

## Calculating confidence intervals for the fixed effects via Wald, bootstrap and profile likelihood
```{r}
confint(early.lmer1,par=5:8,method="Wald",oldNames = FALSE) # Only for fixed effects vc will return NA
confint(early.lmer1,method="boot",boot.type ="perc",oldNames = FALSE,nsim=500)
confint(early.lmer1, level = 0.95,method="profile",oldNames = FALSE)
```

## Get the KR-approximated degrees of freedom
```{r}
#early.lmer1.df.KR = get_ddf_Lb(early.lmer1, fixef(early.lmer1))
```

## Likelihood ratio tests

```{r}
early.lmer1.noprog<-lmer(cog~1+age0+(1 + age0|id), REML = FALSE, data=early.int1)
early.lmer1.intprog<-lmer(cog~1+age0+program+(1 + age0|id), REML = FALSE, data=early.int1)
anova(early.lmer1.noprog,early.lmer1.intprog,early.lmer1)
```

## Random effects covariance matrix
```{r}
D.early = unclass(VarCorr(early.lmer1))$id
D.early
```

## Predicted random effects
```{r}
early.lmer1.re = ranef(early.lmer1)$id
head(early.lmer1.re,10)

plot(early.lmer1.re, main="Random intercept (b0i) versus random slope (b1i)")
```
# OLS vs LM Estimates
## Creating the subject specific intercepts and slopes
```{r}
ind.coef=coef(early.lmer1)$id
head(ind.coef)


int.subject=ind.coef[,1]+ind.coef[,3]
slope.subject=ind.coef[,2]+ind.coef[,4]
plot(int.subject,slope.subject, main="Random intercept versus random slope (Including the fixed effects)")
```

